Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-03-16 paper/abstract.tex
small edits from a big re-read

The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
an age proxy, to reveal their rotational evolution and demonstrate that
kinematics could be a useful tool for calibrating gyrochronology.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
they are young, at old ages we find that late-G and early-K dwarfs rotate at
the {\it same rate} or faster than late-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that the surface rotation rates of K dwarfs go through an epoch
of inhibited evolution.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
K-dwarfs are likely to be synchronized binaries.

>>>>

The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
In this work, we use the velocity dispersions of groups of low-mass \kepler\
dwarfs as an age proxy, to reveal the temperature-dependent rotational
evolution of these stars, and demonstrate that kinematic ages could be a
useful tool for calibrating gyrochronology in future.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
they are young, at old ages we find that late-G and early-K dwarfs rotate at
the {\it same rate} or faster than late-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that the surface rotation rates of K dwarfs go through an epoch
of inhibited evolution.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
K-dwarfs are likely to be synchronized binaries.

### 2020-03-16 paper/intro.tex
small edits from a big re-read

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} confirm that rotational evolution is a highly complex
process.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
sandquist2016},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period--color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.
Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (Ro; the ratio of rotation
period to convective turnover time) of around 2 \citep{vansaders2016,
vansaders2018}.

@@

\subsection{Using kinematics as an age proxy}

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the Galaxy with a low out-of-plane, or vertical, velocity (\vz),
just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
On average, the vertical velocities of older stars is observed to be larger
\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
casagrande2011}.
This is likely either a signature of dynamical heating, such as from
interactions with giant molecular clouds, spiral arms and the galactic bar
\citep[see][for a review of secular evolution in the MW]{sellwood2014}, or an
indication that stars formed dynamically ``hotter'' in the past
\citep[e.g.,][]{bird2013}.
In either case, the vertical velocity distribution is observed to depend
significantly on stellar age.
While the velocity of any individual star only provides a weak age constraint,
because its velocity depends on its current position in its orbit, the
velocity {\it dispersion} of a {\it group} of stars indicates whether that
group is old or young relative to other groups.
% Vertical velocity dispersion is thought to increase monotonically with age.
In this work, we compare the velocity dispersions of groups of field stars in
the Galactic thin disk to ascertain which groups are older and which younger,
and draw conclusions about the rotational evolution of stars based on their
implied relative ages.

>>>>

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and slowly lose angular momentum via magnetic braking
\citep[\eg][]{schatzman1962, weber1967, kraft1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are born with random rotation periods, between 1 and 10 days,
observations of young open clusters reveal that their rotation periods
converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology, the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} confirm that rotational evolution is a highly complex
process.
For example, the early-to-mid M dwarfs in the $\sim$ 650 Myr Praesepe cluster
spin more slowly than the G dwarfs; in theory because lower-mass stars have
deeper convective zones which generate stronger magnetic fields and more
efficient magnetic braking.
However, in the NGC 6811 cluster which is around 1 Gyr \citep{janes2011,
sandquist2016},
late-K dwarfs rotate at the {\it same rate} as early-K dwarfs
\citep{curtis2019}.
% In other words, convection zone depth cannot be the only variable that affects
% stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period--color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.

Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (Ro; the ratio of rotation
period to convective turnover time) of around 2 \citep{vansaders2016,
vansaders2018}.

@@

\subsection{Using velocity dispersion as an age proxy}

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the Galaxy with a low out-of-plane, or vertical, velocity (\vz),
just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
On average, the vertical velocities of older stars is observed to be larger
\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
casagrande2011}.
This is likely either a signature of dynamical heating, such as from
interactions with giant molecular clouds, spiral arms and the galactic bar
\citep[see][for a review of secular evolution in the MW]{sellwood2014}, or an
indication that stars formed dynamically ``hotter'' in the past
\citep[e.g.,][]{bird2013}.
In either case, the vertical velocity distribution is observed to depend
significantly on stellar age.
While the velocity of any individual star only provides a weak age constrain
because its velocity depends on its current position in its orbit, the
velocity {\it dispersion} of a {\it group} of stars indicates whether that
group is old or young relative to other groups.
% Vertical velocity dispersion is thought to increase monotonically with age.
In this work, we compare the velocity dispersions of groups of field stars in
the Galactic thin disk to ascertain which groups are older and which younger,
and draw conclusions about the rotational evolution of stars based on their
implied relative ages.

### 2020-03-16 paper/method.tex
small edits from a big re-read

The rotation periods of the dwarf stars in the \mct\ sample are shown on a
\gaia\ color-magnitude diagram (CMD) in the top panel of figure
\ref{fig:age_gradient}.
In the bottom panel, the stars are colored by their gyrochronal age,
calculated using the \citet{angus2019} gyrochronology relation.
The stars with old gyrochronal ages, plotted in purple hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
The lack of gyrochronologically old M dwarfs suggests that either old M dwarfs
are missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2008, newton2017, kiman2019}, and are therefore more likely
to have measurable rotation periods at old ages, the latter scenario seems
likely.
However, it is also possible that the rotation periods of the oldest early M
dwarfs are so long that they are not measurable with Kepler data.
Ground-based rotation period measurements of mid and late M dwarfs indicate
that there is an upper limit to the rotation periods of {\it late} M dwarfs of
around 140 days \citep{newton2016, newton2018}, which is much
longer than the longest rotation periods measured in the \mct\ sample (around
70 days).
The apparent lack of old gyro-ages for M dwarfs in figure
\ref{fig:age_gradient} may be caused by a combination of ages being
underestimated by a poorly calibrated model, and rotation period detection
bias.
The \citet{angus2019} gyrochronology relation is a simple polynomial model,
fit to the period-color relation of Praesepe.
Inaccuracies at low masses are a typical feature of empirically calibrated
gyrochronology models: since there are no (or at least very few) old M dwarfs
with rotation periods, the models are poorly calibrated for these stars.
\begin{figure}
  \caption{
      Top: de-reddened MS \kepler\ stars with \mct\ rotation periods, plotted
    on a \gaia\ CMD.
    We removed photometric binaries and subgiants from the sample by excluding
    stars above the dashed lines.
    Bottom: a zoom-in of the top panel, with stars colored by their
    gyrochronal age \citep{angus2019}, instead of their rotation period.
    A general age gradient is visible across the main sequence.
    Since the \citet{angus2019} relation predicts that the oldest stars in
    the \mct\ sample are late-G and early-K dwarfs, it is probably
    under-predicting the ages of late-K and early-M dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts_double}
\label{fig:age_gradient}
\end{figure}

>>>>

The rotation periods of the dwarf stars in the \mct\ sample are shown on a
\gaia\ color-magnitude diagram (CMD) in the top panel of figure
\ref{fig:age_gradient}.
In the bottom panel, the stars are colored by their gyrochronal age,
calculated using the \citet{angus2019} gyrochronology relation.
The stars with old gyrochronal ages, plotted in purple hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
The lack of gyrochronologically old M dwarfs suggests that either old M dwarfs
are missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2008, newton2017, kiman2019}, and are therefore more likely
to have measurable rotation periods at old ages, the latter scenario seems
likely.
However, it is also possible that the rotation periods of the oldest early M
dwarfs are so long that they are not measurable with Kepler data.
Ground-based rotation period measurements of mid and late M dwarfs indicate
that there is an upper limit to the rotation periods of {\it late} M dwarfs of
around 140 days \citep{newton2016, newton2018}, which is much
longer than the longest rotation periods measured in the \mct\ sample (around
70 days).
The apparent lack of old gyro-ages for M dwarfs in figure
\ref{fig:age_gradient} may be caused by a combination of ages being
underestimated by a poorly calibrated model, and rotation period detection
bias.
The \citet{angus2019} gyrochronology relation is a simple polynomial model,
fit to the period-color relation of Praesepe.
Inaccuracies at low masses are a typical feature of empirically calibrated
gyrochronology models since there are no (or at least very few) old M dwarfs
with rotation periods and the models are poorly calibrated for these stars.
\begin{figure}
  \caption{
      Top: de-reddened MS \kepler\ stars with \mct\ rotation periods, plotted
    on a \gaia\ CMD.
    We removed photometric binaries and subgiants from the sample by excluding
    stars above the dashed lines.
    Bottom: a zoom-in of the top panel, with stars colored by their
    gyrochronal age \citep{angus2019}, instead of their rotation period.
    A general age gradient is visible across the main sequence.
    Since the \citet{angus2019} relation predicts that the oldest stars in
    the \mct\ sample are late-G and early-K dwarfs, it is probably
    under-predicting the ages of late-K and early-M dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts_double}
\label{fig:age_gradient}
\end{figure}
